\documentclass[11pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage[margin=2.5cm]{geometry}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{fancyhdr}
\usepackage{enumitem}
\usepackage{tabularx}
\usepackage{listings}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{float}
\usepackage{titlesec}
\usepackage{mdframed}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,shapes.geometric}

\definecolor{navy}{HTML}{1A3A8F}
\definecolor{dgray}{HTML}{6B7A99}
\definecolor{codebg}{HTML}{F4F6FA}
\definecolor{codeframe}{HTML}{C8D0E0}
\definecolor{green}{HTML}{2E8B57}
\definecolor{orange}{HTML}{DD6B20}

\lstset{language=Python,basicstyle=\ttfamily\small,
  backgroundcolor=\color{codebg},frame=single,rulecolor=\color{codeframe},
  breaklines=true,keywordstyle=\bfseries\color{navy},
  commentstyle=\itshape\color{dgray},showstringspaces=false,
  numbers=left,numberstyle=\tiny\color{dgray},numbersep=5pt,
  xleftmargin=15pt,xrightmargin=5pt,aboveskip=8pt,belowskip=8pt}

\titleformat{\section}{\large\bfseries\color{navy}}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries\color{navy}}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}{\small\bfseries\itshape\color{dgray}}{\thesubsubsection}{0.5em}{}

\pagestyle{fancy}\fancyhf{}
\fancyhead[L]{\small\color{dgray}psgscoring v0.2.5 --- Handbook}
\fancyhead[R]{\small\color{dgray}psgscoring}
\fancyfoot[C]{\small\color{dgray}\thepage}
\renewcommand{\headrulewidth}{0.4pt}

\newmdenv[backgroundcolor=codebg,linecolor=codeframe,linewidth=1pt,
  innertopmargin=6pt,innerbottommargin=6pt]{infobox}

\title{\textbf{\Large psgscoring}\\[6pt]
\large A Technical Handbook for Automated\\
Polysomnography Scoring\\[8pt]
\normalsize Version 0.2.5 --- April 2026}



\author{psgscoring contributors\\[4pt]
\small\url{https://github.com/bartromb/psgscoring}}

\date{}

\begin{document}
\maketitle
\thispagestyle{fancy}

\begin{abstract}
\noindent
This handbook provides a technical introduction to \texttt{psgscoring},
an open-source Python library (BSD-3) for AASM~2.6-compliant automated
polysomnography (PSG) scoring. It is written for BSc-level computer
science students with basic knowledge of Python, NumPy, and signal
processing. The handbook covers the clinical context, signal processing
pipeline, respiratory event detection algorithms, apnea type
classification, and the architecture of the correction mechanisms that
distinguish \texttt{psgscoring} from na\"ive threshold-based approaches.

\medskip\noindent
\textbf{Prerequisites:} Python 3.9+, NumPy, SciPy, basic DSP
(Fourier transform, filtering, envelopes).\\
\textbf{Repository:} \url{https://github.com/bartromb/psgscoring}\\
\textbf{Installation:} \texttt{pip install psgscoring}
\end{abstract}

\tableofcontents
\newpage

% ================================================================
\section{Clinical Context: Why This Matters}
% ================================================================

\subsection{Sleep Apnea}

Obstructive Sleep Apnea (OSA) affects approximately 936~million adults
worldwide. During sleep, the upper airway collapses partially or
completely, causing breathing interruptions that fragment sleep and reduce
blood oxygen levels. Left untreated, OSA increases the risk of
hypertension, stroke, heart failure, and traffic accidents.

Diagnosis relies on \textbf{polysomnography} (PSG): an overnight
recording of brain waves (EEG), eye movements (EOG), muscle tone (EMG),
airflow, respiratory effort, blood oxygen saturation (SpO$_2$), body
position, snoring, heart rhythm (ECG), and leg movements.

\subsection{The Scoring Problem}

A sleep technologist manually reviews the recording and counts:
\begin{itemize}[nosep]
  \item \textbf{Apneas}: $\geq$90\% flow reduction for $\geq$10~seconds
  \item \textbf{Hypopneas}: $\geq$30\% flow reduction for $\geq$10~seconds
        with $\geq$3\% oxygen desaturation or an EEG arousal
\end{itemize}

The \textbf{Apnea-Hypopnea Index (AHI)} = (apneas + hypopneas) / hours
of sleep, is the primary diagnostic metric:

\begin{center}
\begin{tabular}{lll}
\toprule
\textbf{AHI} & \textbf{Severity} & \textbf{Treatment} \\
\midrule
$<$5     & Normal   & None \\
5--15    & Mild     & Positional therapy, MAD \\
15--30   & Moderate & CPAP \\
$>$30    & Severe   & CPAP (urgent) \\
\bottomrule
\end{tabular}
\end{center}

Manual scoring takes 1--2 hours per study with 10--20\% inter-scorer
variability. \texttt{psgscoring} automates this process.

\subsection{The AASM Standard}

The American Academy of Sleep Medicine (AASM) publishes the
\emph{Manual for the Scoring of Sleep and Associated Events} (current:
Version 2.6, 2020). This is the rulebook. Every threshold, every
duration criterion, every exception in \texttt{psgscoring} traces back
to a specific AASM rule. Key rules:

\begin{itemize}[nosep]
  \item \textbf{Rule 1A}: Hypopnea requires $\geq$3\% SpO$_2$ desaturation
  \item \textbf{Rule 1B}: Hypopnea can alternatively be scored with an
        EEG arousal (without desaturation)
  \item Apneas are scored on the \textbf{thermistor} (oronasal thermal sensor)
  \item Hypopneas are scored on the \textbf{nasal pressure transducer}
\end{itemize}

% ================================================================
\section{Architecture Overview}
% ================================================================

\texttt{psgscoring} is organised as 12 independently importable
submodules with a strict one-directional dependency graph:

\begin{figure}[H]
\centering
\begin{tikzpicture}[
  box/.style={draw,rounded corners=3pt,fill=codebg,
              minimum width=2.3cm,minimum height=0.6cm,
              font=\ttfamily\scriptsize,align=center,inner sep=3pt},
  agg/.style={box,fill=navy!8,minimum width=2.6cm},
  arr/.style={-{Stealth[length=4pt]},color=navy,semithick},
]
% ── Row 1 (y=0): Foundation layer ──
\node[box] (const) at (0,0)   {constants};
\node[box] (utils) at (3.5,0) {utils};
\node[box] (signal) at (7,0)  {signal};

% ── Row 2 (y=-1.8): Signal consumers ──
\node[box] (breath) at (0,-1.8)     {breath};
\node[box] (classify) at (3.5,-1.8) {classify};
\node[box] (spo2) at (7,-1.8)       {spo2};

% ── Row 3 (y=-3.6): Independent modules ──
% ecg_effort shifted to x=5.25 to clear the classify→respiratory column
\node[box] (plm) at (0,-3.6)        {plm};
\node[box] (ecg) at (5.25,-3.6)     {ecg\_effort};
\node[box] (ancil) at (8.5,-3.6)    {ancillary};
\node[box] (sq) at (11.5,-3.6)      {signal\_quality};

% ── Row 4 (y=-5.4): Aggregator ──
\node[agg] (resp) at (3.5,-5.4) {respiratory};

% ── Row 5 (y=-7.0): Master ──
\node[agg] (pipe) at (5.25,-7.0) {pipeline};

% ══ ARROWS ══════════════════════════════════════════════

% constants → utils (straight right)
\draw[arr] (const) -- (utils);

% constants → signal (route above)
\draw[arr,rounded corners=4pt] (const.north) -- ++(0,0.4) -| (signal.north);

% signal → breath (left arc)
\draw[arr,rounded corners=4pt] ([xshift=-0.5cm]signal.south) -- ++(0,-0.45) -| (breath.north);

% signal → classify
\draw[arr,rounded corners=4pt] ([xshift=-0.1cm]signal.south) -- ++(0,-0.3) -| (classify.north);

% signal → spo2 (straight down)
\draw[arr] ([xshift=0.3cm]signal.south) -- ([xshift=0.3cm]spo2.north);

% breath → respiratory (down left column, then right — passes left of plm)
\draw[arr,rounded corners=4pt] (breath.south) -- ++(0,-2.6) -| ([xshift=-0.5cm]resp.north);

% classify → respiratory (straight down — column is now CLEAR)
\draw[arr] (classify) -- (resp);

% spo2 → respiratory (down, then left)
\draw[arr,rounded corners=4pt] (spo2.south) -- ++(0,-2.6) -| ([xshift=0.5cm]resp.north);

% ecg_effort → respiratory (short diagonal)
\draw[arr,rounded corners=4pt] (ecg.south) -- ++(0,-0.5) -| ([xshift=0.8cm]resp.north);

% respiratory → pipeline
\draw[arr,rounded corners=4pt] (resp.south) -- ++(0,-0.5) -| (pipe.north);

% plm → pipeline
\draw[arr,rounded corners=4pt] (plm.south) -- ++(0,-2.0) -| ([xshift=-1.0cm]pipe.north);

% ancillary → pipeline
\draw[arr,rounded corners=4pt] (ancil.south) -- ++(0,-2.0) -| ([xshift=0.8cm]pipe.north);

% signal_quality → pipeline
\draw[arr,rounded corners=4pt] (sq.south) -- ++(0,-2.0) -| ([xshift=1.2cm]pipe.north east);

% ── Row labels ──
\node[font=\tiny\itshape,color=dgray,anchor=east] at (-1.5,0) {foundation};
\node[font=\tiny\itshape,color=dgray,anchor=east] at (-1.5,-1.8) {signal processing};
\node[font=\tiny\itshape,color=dgray,anchor=east] at (-1.5,-3.6) {independent};
\node[font=\tiny\itshape,color=dgray,anchor=east] at (-1.5,-5.4) {aggregation};
\node[font=\tiny\itshape,color=dgray,anchor=east] at (-1.5,-7.0) {orchestrator};

\end{tikzpicture}
\caption{Module dependency graph. Arrows point from dependency to
  dependent module. The five layers correspond to increasing
  abstraction: foundation $\to$ signal processing $\to$ independent
  modules $\to$ aggregation $\to$ orchestrator.}
\end{figure}

\begin{table}[H]
\centering\small
\caption{Submodule responsibilities and approximate size}
\begin{tabular}{@{}llr@{}}
\toprule
\textbf{Module} & \textbf{Responsibility} & \textbf{LOC} \\
\midrule
\texttt{constants} & AASM thresholds, band limits, scoring profiles & 76 \\
\texttt{utils}     & Sleep mask construction, channel detection & 134 \\
\texttt{signal}    & Linearisation, baseline, MMSD validation & 327 \\
\texttt{breath}    & Breath segmentation, flattening index & 254 \\
\texttt{classify}  & 7-rule apnea type decision tree & 310 \\
\texttt{spo2}      & SpO$_2$ coupling, ODI computation & 218 \\
\texttt{ecg\_effort} & ECG-derived effort (TECG + spectral) & 180 \\
\texttt{plm}       & Periodic Limb Movement detection & 271 \\
\texttt{ancillary} & Heart rate, snoring, position, CSR & 277 \\
\texttt{signal\_quality} & Per-channel quality grading & 160 \\
\texttt{respiratory} & Apnea/hypopnea detection + 12 corrections & 1061 \\
\texttt{pipeline}  & Master orchestrator & 420 \\
\bottomrule
\end{tabular}
\end{table}

% ================================================================
\section{Signal Processing Chain}
\label{sec:signal}
% ================================================================

The respiratory analysis begins with transforming the raw sensor data
into a form suitable for event detection. This section describes each
step mathematically and in code.

\subsection{Step 1: Nasal Pressure Linearisation}

The nasal pressure transducer measures \emph{pressure}, not flow. The
relationship between pressure and flow follows Bernoulli's equation:

\begin{equation}
  \text{flow} \propto \sqrt{|\text{pressure}|}
\end{equation}

The inverse square-root transform linearises the signal:

\begin{equation}
  x_{\text{lin}}(t) = \text{sign}(x(t)) \cdot \sqrt{|x(t)|}
\end{equation}

\begin{lstlisting}
def linearise_nasal_pressure(signal):
    """Bernoulli square-root linearisation."""
    return np.sign(signal) * np.sqrt(np.abs(signal))
\end{lstlisting}

\begin{infobox}
\textbf{Why this matters:} Without linearisation, the amplitude of
hypopneas is systematically underestimated, because the pressure
signal compresses peaks relative to the true flow.
\end{infobox}

\subsection{Step 2: Bandpass Filtering}

A 4th-order Butterworth bandpass filter (0.05--3~Hz) removes:
\begin{itemize}[nosep]
  \item DC drift and very slow baseline wander ($<$0.05~Hz)
  \item High-frequency noise, cardiac artefact ($>$3~Hz)
\end{itemize}

\begin{lstlisting}
from scipy.signal import butter, sosfiltfilt

def bandpass_flow(signal, sf, low=0.05, high=3.0, order=4):
    sos = butter(order, [low, high], btype='band',
                 fs=sf, output='sos')
    return sosfiltfilt(sos, signal)
\end{lstlisting}

The normal respiratory rate is 12--20 breaths/minute (0.2--0.33~Hz),
well within this passband.

\subsection{Step 3: Hilbert Envelope}

The instantaneous amplitude envelope is extracted via the analytic
signal:

\begin{equation}
  A(t) = |x(t) + i\mathcal{H}\{x(t)\}|
\end{equation}

where $\mathcal{H}$ is the Hilbert transform. This provides a smooth,
non-negative representation of breath amplitude.

\begin{lstlisting}
from scipy.signal import hilbert

def compute_envelope(signal):
    analytic = hilbert(signal)
    return np.abs(analytic)
\end{lstlisting}

\begin{infobox}
\textbf{Key insight:} The envelope ``smooths over'' individual breath
cycles. This is useful for detecting apneas (amplitude $\to$ 0) but
can miss hypopneas where individual breath \emph{peaks} are reduced
but the envelope remains above threshold. This limitation motivates
the peak-based detection added later (Section~\ref{sec:undercounting}).
\end{infobox}

\subsection{Step 4: Dynamic Baseline}

The patient's ``normal'' breathing amplitude varies throughout the
night (deeper in N3, shallower in REM, variable with position changes).
A static threshold would fail. Instead, we compute a dynamic baseline:

\begin{equation}
  B(t) = P_{95}\big(A(\tau) : \tau \in [t - 150\text{s}, t + 150\text{s}]\big)
\end{equation}

A 5-minute sliding window, 95th percentile, advancing in 10-second
steps with linear interpolation between anchor points.

\begin{lstlisting}
def compute_dynamic_baseline(envelope, sf,
                              window_s=300, step_s=10):
    n = len(envelope)
    anchors = range(0, n, int(step_s * sf))
    half_w = int(window_s * sf / 2)
    bl = np.zeros(n)
    for a in anchors:
        s = max(0, a - half_w)
        e = min(n, a + half_w)
        bl[a] = np.percentile(envelope[s:e], 95)
    # Interpolate between anchors
    return np.interp(np.arange(n),
                     list(anchors), bl[list(anchors)])
\end{lstlisting}

\subsection{Step 5: Event Detection (Threshold Crossing)}

An event is detected when the normalised amplitude drops below a
threshold for $\geq$10 seconds:

\begin{equation}
  \frac{A(t)}{B(t)} \leq 1 - \theta \quad \text{for } \geq 10\text{s}
\end{equation}

where $\theta = 0.90$ for apneas (thermistor) and $\theta = 0.30$ for
hypopneas (nasal pressure). The normalised signal is:

\begin{equation}
  \hat{x}(t) = \frac{A(t)}{B(t)}
\end{equation}

Connected components below the threshold are found using
\texttt{scipy.ndimage.label()}.

\subsection{Step 6: MMSD Validation}

Not every flat segment is a true apnea---some are artefacts (electrode
disconnection). The Mean Magnitude of Second Derivative distinguishes
them:

\begin{equation}
  \text{MMSD} = \frac{1}{N}\sum_{i=2}^{N}|x''(i)|
\end{equation}

True apneas have low but non-zero MMSD (residual cardiac oscillation);
artefacts have MMSD $\approx$ 0 (identically flat).

% ================================================================
\section{Apnea Type Classification}
\label{sec:classify}
% ================================================================

After detecting an event, we must determine its type: \textbf{obstructive}
(airway blocked despite effort), \textbf{central} (no respiratory
effort), or \textbf{mixed} (starts central, ends obstructive).

This is a clinical decision that depends on the thoraco-abdominal
effort signals (RIP bands). A 7-rule decision tree processes the event:

\subsection{Feature Extraction}

For each event, the following features are computed from the thorax
and abdomen RIP channels:

\begin{table}[H]
\centering\small
\begin{tabular}{@{}lp{8cm}@{}}
\toprule
\textbf{Feature} & \textbf{Computation} \\
\midrule
\texttt{effort\_ratio} & RMS(effort during event) / RMS(baseline) \\
\texttt{raw\_var\_ratio} & Variance(raw signal during event) / Variance(baseline) \\
\texttt{paradox\_corr} & Pearson correlation(thorax, abdomen) --- negative = paradoxical \\
\texttt{phase\_angle\_deg} & Mean $|\Delta\phi|$ via Hilbert phase of each RIP channel \\
\texttt{quarter\_efforts} & Effort ratio for each quarter of the event duration \\
\texttt{first\_ratio, second\_ratio} & Effort ratio for first vs.\ second half \\
\texttt{flattening\_index} & Mean inspiratory flattening of overlapping breaths \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Decision Tree}

\begin{table}[H]
\centering\small
\caption{Classification rules (evaluated in order; first match wins)}
\begin{tabular}{@{}cllc@{}}
\toprule
\textbf{Rule} & \textbf{Condition} & \textbf{Type} & \textbf{Confidence} \\
\midrule
0 & Phase angle $\geq 45^\circ$ & Obstructive & 0.75--0.97 \\
1 & Paradoxical + high raw variability & Obstructive & 0.85--0.95 \\
2 & High variability, low envelope & Obstructive & 0.70--0.85 \\
3 & First half low effort, second half high & Mixed & 0.70--0.95 \\
4 & Clear effort present & Obstructive & 0.65--0.80 \\
5 & Truly flat, no effort, no paradox & Central & 0.70--0.90 \\
5a & Low effort (gray zone), no obstructive evidence & Central & 0.45--0.75 \\
5b & ECG confirms no inspiratory effort (TECG) & Central & 0.75--0.85 \\
6 & Low effort default $\to$ central; else $\to$ obstructive & varies & $<$0.60 \\
\bottomrule
\end{tabular}
\end{table}

\begin{infobox}
\textbf{The cardiac pulsation problem:} During a true central apnea,
the heart continues to beat. These mechanical oscillations appear on
the thorax/abdomen RIP bands as low-amplitude periodic signals. A
na\"ive algorithm interprets this as ``effort present'' $\to$
obstructive. Rules 5, 5a, and 5b are specifically designed to handle
this, using relaxed thresholds and ECG-derived effort analysis.
\end{infobox}

\subsection{Code Example}

\begin{lstlisting}
from psgscoring import classify_apnea_type

typ, confidence, detail = classify_apnea_type(
    onset_idx=start_sample,
    end_idx=end_sample,
    thorax_env=thorax_envelope,
    abdomen_env=abdomen_envelope,
    thorax_raw=thorax_raw_signal,
    abdomen_raw=abdomen_raw_signal,
    effort_baseline=effort_bl,
    sf=256,
    flattening_index=0.35,  # from breath analysis
)
# typ: "obstructive", "central", or "mixed"
# confidence: 0.0 - 1.0
# detail: {"rule_index": 0, "decision_reason": "..."}
\end{lstlisting}

% ================================================================
\section{SpO$_2$ Coupling}
% ================================================================

AASM Rule 1A requires a $\geq$3\% oxygen desaturation associated with
each hypopnea. The coupling algorithm:

\begin{enumerate}[nosep]
  \item Compute the SpO$_2$ baseline: 90th percentile during sleep
  \item For each candidate hypopnea, search for the nadir in a
        \textbf{post-event window} (configurable: 30--45~seconds)
  \item Desaturation = pre-event SpO$_2$ $-$ nadir
  \item If desaturation $\geq$ 3\%: hypopnea confirmed (Rule 1A)
\end{enumerate}

\begin{infobox}
\textbf{Circulatory delay:} The finger pulse oximeter has a 20--40
second delay relative to the lungs. A 30-second window misses nadirs
arriving at 30--45~seconds. The \emph{standard} scoring profile uses
45~seconds; \emph{strict} uses 30~seconds.
\end{infobox}

% ================================================================
\section{Over-Counting Corrections}
\label{sec:overcounting}
% ================================================================

Na\"ive threshold-based scoring systematically \emph{inflates} AHI.
Six mechanisms are identified and corrected:

\subsection{Fix 1: Post-Apnea Baseline Inflation}

After an apnea, the patient hyperventilates (compensatory breathing).
These large recovery breaths enter the 5-minute baseline window,
inflating it. Subsequent normal breaths then appear ``reduced''
relative to the inflated baseline $\to$ false hypopneas.

\textbf{Solution:} Exclude a 30-second recovery mask after each apnea
from the baseline computation.

\begin{lstlisting}
# Pseudocode
for apnea in detected_apneas:
    mask[apnea.end : apnea.end + 30*sf] = True
baseline = percentile_95(envelope[~mask], window=300s)
\end{lstlisting}

\subsection{Fix 2: SpO$_2$ Cross-Contamination}

At AHI $>$ 60/h, events follow each other so quickly that the SpO$_2$
nadir of event $N$ falls within the search window of event $N+1$.
Event $N+1$ gets an undeserved desaturation $\to$ false hypopnea.

\textbf{Solution:} Flag events where the preceding event's nadir
window overlaps. Since v0.8.14, this is informational only (desaturation
is always computed).

\subsection{Fix 3: Cheyne-Stokes Flagging}

Cheyne-Stokes respiration (CSR) produces a crescendo-decrescendo
pattern. Each trough can cross the 30\% threshold $\to$ scored as
hypopnea. But these are not independent events; they are one central
breathing pattern.

\textbf{Solution:} Detect periodicity via autocorrelation. Events
whose inter-event interval matches the CSR cycle ($\pm$12~s) are
flagged as \texttt{csr\_flagged}.

\subsection{Fix 4: Confidence Stratification}

When RIP signal quality is poor, Rule~6 defaults to ``obstructive.''
This inflates OAHI.

\textbf{Solution:} Report events at four confidence tiers
($\geq$0.85, $\geq$0.60, $\geq$0.40, all). The PDF report shows
OAHI at each threshold.

\subsection{Fix 5: Artefact-Flank Exclusion}

After a flat-line gap (electrode disconnection), the recovery ramp
can cross the flow-reduction threshold.

\textbf{Solution:} Append a 15-second exclusion window after each
flat-line segment $\geq$10~seconds.

\subsection{Fix 6: Local Baseline Validation}

The rolling baseline can be inflated by recovery hyperpnea. Normal
breathing then appears reduced $\to$ false-positive hypopneas of
60--80+ seconds.

\textbf{Solution:} Compare the event amplitude with the 30-second
pre-event window. If the local reduction is $<$20\%, the event is
rejected.

\begin{lstlisting}
def _validate_local_reduction(envelope, onset, end, sf):
    pre_start = max(0, onset - int(30 * sf))
    pre_mean = np.mean(envelope[pre_start:onset])
    event_mean = np.mean(envelope[onset:end])
    reduction = 1 - event_mean / pre_mean
    return reduction >= 0.20, reduction
\end{lstlisting}

% ================================================================
\section{Under-Counting Corrections}
\label{sec:undercounting}
% ================================================================

The complementary problem: the algorithm \emph{misses} events that
a human scorer would detect.

\subsection{Fix 7: Peak-Based Breath Detection}

AASM defines hypopnea as ``peak signal excursions drop $\geq$30\%.''
The Hilbert envelope smooths over breath cycles, missing subtle
per-breath reductions.

\textbf{Solution:} Detect individual breaths via zero-crossings,
compute per-breath peak amplitude, and use a combined mask
(envelope OR peak-based).

\subsection{Fix 8: Extended SpO$_2$ Window}

Finger oximetry has 20--40~s circulatory delay. The 30~s window
misses late nadirs.

\textbf{Solution:} Extend to 45~s (configurable per scoring profile).

\subsection{Fix 9: Flow Smoothing}

Brief oscillations above the 70\% threshold fragment a long event
into sub-10~s segments that are each discarded.

\textbf{Solution:} 3-second rolling mean before thresholding
(configurable: 0/3/5~s per profile).

\subsection{Fix 10: Position Signal Auto-Mapping}

Some EDF systems store position as raw ADC values (0--255) instead
of standard codes (0--4).

\textbf{Solution:} Auto-detect encoding; apply percentile-based
quantisation.

\subsection{Fix 11: Configurable Scoring Profiles}

\begin{table}[H]
\centering\small
\caption{Three predefined scoring profiles}
\begin{tabular}{@{}lccccc@{}}
\toprule
\textbf{Profile} & \textbf{Threshold} & \textbf{SpO$_2$ win.} &
\textbf{Smooth} & \textbf{Peak det.} \\
\midrule
Strict    & $\geq$30\% & 30~s & None & Envelope only \\
Standard  & $\geq$30\% & 45~s & 3~s  & Peak + envelope \\
Sensitive & $\geq$25\% & 45~s & 5~s  & Peak + envelope \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Fix 12: Flattening-Based RERA Detection}

RERAs (Respiratory Effort-Related Arousals) are flow-limited segments
without amplitude reduction, terminated by an arousal. The flattening
index measures the fraction of the inspiratory segment exceeding 80\%
of peak amplitude:

\begin{equation}
  FI = \frac{|\{t : x(t) > 0.8 \cdot x_{\max}\}|}{|\{t : x(t) > 0\}|}
\end{equation}

Normal triangular waveform: $FI \approx 0.10$.
Flattened (flow limitation): $FI > 0.30$.

% ================================================================
\section{ECG-Derived Effort Classification}
% ================================================================

The routinely recorded ECG contains information about respiratory
effort --- if you know where to look.

\subsection{Spectral Effort Classifier}

During a true central apnea, the RIP signal should show only cardiac
pulsation. The spectral classifier computes PSD of the RIP signal
during each event:

\begin{itemize}[nosep]
  \item Cardiac band: 0.8--2.5~Hz (adaptive per patient HR)
  \item Respiratory band: 0.1--0.5~Hz
\end{itemize}

If $>$75\% of power is in the cardiac band and $<$20\% in respiratory
$\to$ flag as probable central.

\subsection{Transformed ECG (TECG)}

Following Berry et al.\ (JCSM 2019):
\begin{enumerate}[nosep]
  \item High-pass filter ECG at 30~Hz
  \item Blank QRS complexes ($\pm$50~ms around R-peaks)
  \item The residual reveals inspiratory EMG bursts from intercostal
        muscles
  \item If no bursts detected $\to$ no respiratory effort $\to$ central
\end{enumerate}

\begin{lstlisting}
from psgscoring.ecg_effort import ecg_effort_assessment

result = ecg_effort_assessment(
    ecg_data, thorax_raw, abdomen_raw,
    sf_ecg=256, sf_effort=256,
    onset_idx=start, end_idx=end,
    r_peaks=r_peak_indices
)
# result["reclassify_as_central"]: bool
# result["ecg_effort_present"]: bool
\end{lstlisting}

% ================================================================
\section{Arousal Detection}
\label{sec:arousals}
% ================================================================

An \textbf{arousal} is a brief awakening from sleep (3--15~seconds of
alpha/beta EEG activity) that fragments sleep architecture without
the patient becoming fully conscious. Arousals are clinically
important because they cause daytime sleepiness even when AHI is low.

\subsection{AASM Arousal Criteria}

The AASM defines an arousal as:
\begin{itemize}[nosep]
  \item An abrupt shift in EEG frequency to alpha, theta, or
        $>$16~Hz (not spindles)
  \item Duration $\geq$3~seconds
  \item Preceded by $\geq$10~seconds of stable sleep
  \item During REM: requires concurrent increase in chin EMG
\end{itemize}

The \textbf{Arousal Index (AI)} = total arousals / hours of sleep.
Normal: $<$15/h. The AI is further broken down:
\begin{itemize}[nosep]
  \item \textbf{Respiratory arousals}: occurring within 3~s of a
        respiratory event termination
  \item \textbf{PLM-associated arousals}: occurring within 0.5~s of a
        leg movement
  \item \textbf{Spontaneous arousals}: all remaining
\end{itemize}

\subsection{Detection Algorithm}

YASAFlaskified implements a two-phase arousal detection in
\texttt{arousal\_analysis.py}:

\textbf{Phase 1 --- EEG-based detection:}
\begin{enumerate}[nosep]
  \item Compute spectral power in the alpha (8--12~Hz) and beta
        ($>$16~Hz) bands using a 1-second sliding window
  \item Detect segments where alpha+beta power exceeds a dynamic
        threshold (mean + 2$\times$SD of the surrounding 30-second epoch)
  \item Merge segments separated by $<$1~s; discard segments $<$3~s
  \item During REM: require concurrent EMG amplitude increase
        ($>$2$\times$ baseline chin EMG RMS)
\end{enumerate}

\textbf{Phase 2 --- Filtering and coupling:}
\begin{enumerate}[nosep]
  \item \textbf{K-complex exclusion}: if the first 1~s of a candidate
        contains a negative peak $<$$-75$~$\mu$V followed by a positive
        peak $>$30~$\mu$V, the minimum duration is raised from 3~s to
        5~s (prevents scoring the trailing alpha rebound as a standalone
        arousal)
  \item \textbf{CVR confidence boost}: if the ECG shows a
        bradycardia during the event followed by sympathetic tachycardia
        ($\geq$10~bpm rise within 15~s), a confidence boost of
        0.10--0.20 is added
  \item \textbf{Respiratory coupling}: arousals within 3~s of a
        respiratory event termination are tagged as respiratory arousals
  \item \textbf{Rule 1B reinstatement}: flow reductions with an
        arousal but without $\geq$3\% desaturation are reinstated as
        hypopneas (AASM Rule~1B)
\end{enumerate}

\begin{infobox}
\textbf{Why K-complex exclusion matters:} K-complexes are normal NREM
phenomena that produce a transient power increase in arousal-detection
bands. Without this filter, up to 15\% of detected ``arousals'' are
actually K-complexes --- inflating the arousal index and triggering
false Rule~1B reinstatements.
\end{infobox}

\subsection{RERA Detection}

RERAs (Respiratory Effort-Related Arousals) are detected from two
sources:
\begin{enumerate}[nosep]
  \item \textbf{FRI-RERA}: flow reductions ($\geq$30\%, $\geq$10~s)
        without desaturation, coupled to an arousal within 15~s of
        event termination
  \item \textbf{Flattening-RERA}: sequences of $\geq$3 consecutive
        breaths with flattening index $>$0.30 spanning $\geq$10~s,
        terminated by an arousal, \emph{without} meeting hypopnea
        amplitude criteria
\end{enumerate}

\begin{equation}
  \text{RDI} = \text{AHI} + \frac{n_{\text{FRI-RERA}} + n_{\text{flat-RERA}}}{\text{TST (hours)}}
\end{equation}

The RDI is clinically relevant for Upper Airway Resistance Syndrome
(UARS), where AHI may be normal but the patient presents with
excessive daytime sleepiness.

% ================================================================
\section{Snoring Analysis}
% ================================================================

Snoring is detected from the microphone or vibration sensor channel
(typically labelled ``Snore'' or ``Phono'' in the EDF).

\subsection{Detection Method}

\begin{enumerate}[nosep]
  \item Bandpass filter the snore channel (30--300~Hz) to isolate
        snoring frequencies
  \item Compute the RMS amplitude in 1-second windows
  \item Determine a dynamic snore threshold: the 60th percentile of
        the RMS amplitude during sleep epochs (P60)
  \item Segments where RMS $>$ P60 for $\geq$3~consecutive seconds
        are classified as snoring epochs
\end{enumerate}

\subsection{Output Metrics}

\begin{table}[H]
\centering\small
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{Metric} & \textbf{Definition} \\
\midrule
Snore duration (min) & Total minutes of detected snoring \\
Snore \% of TST & Fraction of sleep time with snoring \\
Snore index (/h) & Snoring episodes per hour of sleep \\
\bottomrule
\end{tabular}
\end{table}

\begin{lstlisting}
# Access snoring results
snore = results["snore"]["summary"]
print(f"Snoring: {snore['snore_pct_tst']:.1f}% of sleep")
print(f"Snore index: {snore['snore_index']:.1f}/h")
\end{lstlisting}

\begin{infobox}
\textbf{Clinical note:} Snoring alone is not diagnostic, but a high
snore index combined with mild AHI may indicate UARS. The position
overlay in the PDF report shows whether snoring is position-dependent
(common: supine-dominant snoring).
\end{infobox}

% ================================================================
\section{Heart Rate Analysis}
% ================================================================

Heart rate (HR) is derived from the ECG channel via R-peak detection.

\subsection{R-Peak Detection}

\begin{enumerate}[nosep]
  \item Bandpass filter ECG (5--30~Hz, 4th-order Butterworth) to
        enhance QRS complexes
  \item Apply \texttt{scipy.signal.find\_peaks()} with adaptive
        height threshold (0.6$\times$ 90th percentile of filtered signal)
        and minimum distance (0.3~s = 200~bpm maximum)
  \item Compute R-R intervals; reject outliers ($<$0.3~s or $>$2.0~s)
  \item Instantaneous HR = 60 / R-R interval (bpm)
\end{enumerate}

\subsection{Output Metrics}

\begin{table}[H]
\centering\small
\begin{tabular}{@{}llp{5.5cm}@{}}
\toprule
\textbf{Metric} & \textbf{Unit} & \textbf{Clinical significance} \\
\midrule
Mean HR & bpm & Normal: 60--100 during sleep \\
Min HR  & bpm & Bradycardia $<$50; check medication (beta-blockers) \\
Max HR  & bpm & Tachycardia $>$100; may indicate autonomic arousal \\
Bradycardia episodes & count & Number of segments $<$50~bpm $\geq$10~s \\
Tachycardia episodes & count & Number of segments $>$100~bpm $\geq$10~s \\
\bottomrule
\end{tabular}
\end{table}

\subsection{CVR Pattern}

The Cyclic Variation of heart Rate (CVR) is a characteristic pattern
in OSA: \emph{bradycardia} during the apnea (vagal tone from
hypoxia), followed by \emph{tachycardia} upon arousal (sympathetic
activation). The magnitude of this swing ($\Delta$HR $\geq$10~bpm
within 15~s) is used as a confidence boost for arousal detection
(Section~\ref{sec:arousals}).

\begin{center}
\begin{tikzpicture}[scale=0.9]
  \draw[->] (0,0) -- (8,0) node[right] {\small time};
  \draw[->] (0,0) -- (0,3) node[above] {\small HR (bpm)};
  % Baseline
  \draw[dashed,dgray] (0,1.5) -- (8,1.5) node[right] {\tiny 70 bpm};
  % CVR pattern
  \draw[thick,navy]
    (0.5,1.5) -- (1,1.5) -- (1.5,0.8) -- (2.5,0.6) -- (3,0.8)
    -- (3.3,2.5) -- (3.8,2.2) -- (4.5,1.5)
    -- (5,1.5) -- (5.5,0.8) -- (6.5,0.6) -- (7,0.8)
    -- (7.3,2.5) -- (7.8,2.2);
  % Labels
  \draw[<->,red,thick] (1.5,0.6) -- (1.5,1.5);
  \node[red,font=\tiny,right] at (1.6,1.0) {brady};
  \draw[<->,green,thick] (3.3,1.5) -- (3.3,2.5);
  \node[green,font=\tiny,right] at (3.4,2.0) {tachy};
  % Apnea markers
  \fill[red!15] (1,0) rectangle (3,-0.3);
  \node[font=\tiny] at (2,-0.15) {apnea};
  \fill[red!15] (5,0) rectangle (7,-0.3);
  \node[font=\tiny] at (6,-0.15) {apnea};
  \draw[<->,orange,thick] (3,-0.5) -- node[below,font=\tiny] {arousal} (3.5,-0.5);
\end{tikzpicture}
\end{center}

% ================================================================
\section{Body Position Analysis}
% ================================================================

Body position during sleep determines whether OSA is
\emph{position-dependent} --- a clinically important distinction
because position-dependent OSA may be treated with positional therapy
(e.g., tennis ball, NightBalance device) instead of CPAP.

\subsection{Position Mapping}

The EDF position channel uses one of three encoding schemes:

\begin{table}[H]
\centering\small
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Encoding} & \textbf{Format} & \textbf{Systems} \\
\midrule
Integer codes & 0=supine, 1=left, 2=prone, 3=right, 4=upright & Standard \\
Raw ADC & 0--255 (voltage proportional to tilt angle) & SomnoMedics, Embla \\
String labels & ``Supine'', ``Left'', etc. & Some Natus systems \\
\bottomrule
\end{tabular}
\end{table}

The auto-mapping function (\texttt{\_map\_position\_signal()}) detects
the encoding and applies percentile-based quantisation for raw ADC
values:

\begin{lstlisting}
# Automatic position mapping
pos = results["position"]["summary"]
for position in ["supine", "left", "right", "prone", "upright"]:
    pct = pos.get(f"pct_{position}", 0)
    ahi = pos.get(f"ahi_{position}", 0)
    print(f"  {position:8s}: {pct:.0f}% of TST, AHI={ahi:.1f}/h")
\end{lstlisting}

\subsection{Positional AHI}

Events are mapped to body position via the position channel at event
onset. The position-specific AHI is:

\begin{equation}
  \text{AHI}_{\text{pos}} = \frac{n_{\text{events in position}}}
    {\text{time in position (hours)}}
\end{equation}

\textbf{Supine-dominant OSA} is defined as supine AHI $\geq$ 2$\times$
non-supine AHI. This phenotype is more prevalent in younger, non-obese
patients and responds well to positional therapy.

\subsection{Position-Induced Baseline Reset}

When a patient changes position (e.g., supine $\to$ lateral), the
respiratory amplitude changes abruptly. The dynamic baseline must
reset to avoid false events at the transition. \texttt{psgscoring}
detects position changes and resets the baseline window at each
transition (v0.8.25).

% ================================================================
\section{Cheyne-Stokes Detection}
% ================================================================

Cheyne-Stokes Respiration (CSR) is a crescendo-decrescendo breathing
pattern characteristic of central sleep apnea, most commonly seen in
heart failure patients.

\subsection{Detection via Autocorrelation}

The flow envelope is analysed using autocorrelation to detect periodic
breathing patterns:

\begin{enumerate}[nosep]
  \item Compute the autocorrelation of the Hilbert envelope over a
        5-minute window
  \item Search for a secondary peak between 40 and 100~seconds
        (typical CSR cycle length: 45--75~seconds)
  \item If the peak-to-trough ratio exceeds a threshold (1.5$\times$
        the surrounding baseline), CSR is detected
  \item Determine the CSR periodicity from the peak location
\end{enumerate}

\begin{equation}
  R(\tau) = \frac{1}{N}\sum_{t=0}^{N-\tau} A(t) \cdot A(t+\tau)
\end{equation}

\subsection{Event Flagging}

After CSR detection, individual respiratory events whose inter-event
interval matches the detected periodicity ($\pm$12~s) are retroactively
flagged:

\begin{lstlisting}
# CSR results
csr = results["respiratory"]["summary"]
print(f"CSR detected: {csr.get('csr_detected', False)}")
print(f"CSR period:   {csr.get('csr_period_s', 0):.0f}s")
print(f"Events flagged: {csr.get('n_csr_flagged', 0)}")
print(f"AHI without CSR: {csr.get('ahi_csr_corrected', 0):.1f}/h")
\end{lstlisting}

\begin{infobox}
\textbf{Clinical impact:} CSR requires fundamentally different treatment
(Adaptive Servo-Ventilation, ASV) rather than standard CPAP. Without
CSR flagging, a heart failure patient may receive an AHI~$>$60/h and
be started on CPAP, which may worsen central apneas. The corrected AHI
(excluding CSR-flagged events) helps the clinician distinguish
obstructive from central pathology.
\end{infobox}

% ================================================================
\section{Sleep Cycles and Architecture}
% ================================================================

While sleep staging is performed by YASA (upstream of
\texttt{psgscoring}), the platform computes additional sleep
architecture metrics.

\subsection{Sleep Cycle Detection}

Sleep cycles are detected following Feinberg \& Floyd (1979) criteria:

\begin{itemize}[nosep]
  \item A NREM period requires $\geq$15~minutes of continuous NREM
        sleep (N1, N2, or N3)
  \item REM consolidation: REM segments separated by $<$2~minutes of
        NREM are merged into a single REM period
  \item A cycle = one NREM period + the subsequent REM period
  \item Normal sleep: 4--6 cycles per night, each 90--120~minutes
\end{itemize}

\subsection{Stage-Specific Metrics}

Sleep architecture affects respiratory scoring because:
\begin{itemize}[nosep]
  \item N3 (deep sleep): respiratory amplitude is lower $\to$ the
        dynamic baseline must be stage-blended
  \item REM: muscle atonia increases upper airway collapsibility $\to$
        REM-AHI is typically higher than NREM-AHI
  \item REM-dominant OSA (REM-AHI/NREM-AHI $>$2) is a distinct
        phenotype, more common in women and younger patients
\end{itemize}

\begin{lstlisting}
# Stage-specific AHI
resp = results["respiratory"]["summary"]
print(f"REM AHI:  {resp.get('ahi_rem', '?')}/h")
print(f"NREM AHI: {resp.get('ahi_nrem', '?')}/h")
ratio = resp.get('ahi_rem', 0) / max(resp.get('ahi_nrem', 1), 1)
if ratio > 2:
    print("REM-dominant OSA phenotype")
\end{lstlisting}

\subsection{Sleep Stage Transition Matrix}

The PDF report includes a 5$\times$5 transition matrix showing the
number of stage transitions (W/N1/N2/N3/R $\to$ W/N1/N2/N3/R). This
helps identify:
\begin{itemize}[nosep]
  \item Excessive W$\to$N1 transitions: fragmented sleep onset
  \item High N2$\to$W count: frequent arousals from light sleep
  \item Low N2$\to$N3 transitions: reduced deep sleep initiation
  \item Absent or rare transitions to R: REM suppression (medication,
        depression, severe OSA)
\end{itemize}

% ================================================================
\section{The Complete Pipeline}
% ================================================================

The master function \texttt{run\_pneumo\_analysis()} orchestrates
all modules:

\begin{lstlisting}
from psgscoring.pipeline import run_pneumo_analysis

results = run_pneumo_analysis(
    raw_pneumo,         # MNE Raw object
    hypno,              # ["W","N1","N2",...]
    channel_map,        # {"flow": "Pressure", ...}
    scoring_profile="standard",
)
# results["respiratory"]["events"]: list of dicts
# results["respiratory"]["summary"]["ahi_total"]
# results["spo2"]["summary"]["odi_3pct"]
# results["plm"]["summary"]["plmi"]
\end{lstlisting}

\begin{figure}[H]
\centering
\begin{tikzpicture}[
  node distance=0.5cm,
  pstep/.style={draw,rounded corners,fill=codebg,minimum width=8cm,
               minimum height=0.5cm,font=\small,align=center},
  arr/.style={-{Stealth[length=4pt]},thick,color=navy}
]
\node[pstep] (s1) {1. EDF load + channel mapping};
\node[pstep,below=of s1] (s2) {2. Signal quality assessment};
\node[pstep,below=of s2] (s3) {3. Linearisation + filtering + envelope};
\node[pstep,below=of s3] (s4) {4. Dynamic baseline computation};
\node[pstep,below=of s4] (s5) {5. Apnea detection (thermistor, $\geq$90\%)};
\node[pstep,below=of s5] (s6) {6. Hypopnea detection (nasal pressure, $\geq$30\%)};
\node[pstep,below=of s6] (s7) {7. SpO$_2$ coupling + type classification};
\node[pstep,below=of s7] (s8) {8. 6 over-counting corrections};
\node[pstep,below=of s8] (s9) {9. 6 under-counting corrections};
\node[pstep,below=of s9] (s10) {10. RERA/RDI + arousal coupling};
\node[pstep,below=of s10] (s11) {11. Summary statistics (AHI, ODI, PLMI)};

\foreach \i/\j in {s1/s2,s2/s3,s3/s4,s4/s5,s5/s6,s6/s7,s7/s8,s8/s9,s9/s10,s10/s11}
  \draw[arr] (\i) -- (\j);
\end{tikzpicture}
\caption{Processing pipeline steps}
\end{figure}

% ================================================================
\section{Testing}
% ================================================================

\texttt{psgscoring} includes 47 tests (Python 3.9--3.12 CI):

\subsection{Unit Tests (37)}

Each module has dedicated tests covering edge cases:
signal processing with zero-length arrays, classification with
extreme feature values, SpO$_2$ coupling at boundary conditions.

\subsection{Regression Tests (7)}

Synthetic signals with known ground truth:
\begin{itemize}[nosep]
  \item Obstructive apnea: effort present + flow absent $\to$ must classify obstructive
  \item Central apnea: no effort + no flow $\to$ must classify central
  \item Mixed apnea: first half central, second half obstructive
  \item Dynamic baseline stability on constant-amplitude input
  \item SpO$_2$ desaturation detection with known nadir
  \item Breath count for 12 breaths/minute signal
  \item Flattening index passthrough to classification detail
\end{itemize}

\subsection{Property-Based Tests (3, Hypothesis)}

Random inputs to \texttt{classify\_apnea\_type()}:
\begin{itemize}[nosep]
  \item 500 random combinations: never crashes, always returns valid type
  \item Low effort inputs: must not produce high-confidence obstructive
  \item Short segments (2--100 samples): crash-free
\end{itemize}

\begin{lstlisting}
# Running tests
cd psgscoring
python -m pytest tests/ -v
\end{lstlisting}

% ================================================================
\section{Data Structures}
% ================================================================

\subsection{Event Dictionary}

Each respiratory event is represented as a Python dictionary:

\begin{lstlisting}
{
    "type":               "obstructive",  # or "central"/"mixed"/"hypopnea"
    "onset_s":            1234.5,         # seconds from recording start
    "duration_s":         18.0,
    "stage":              "N2",           # sleep stage at event onset
    "desaturation_pct":   4.2,            # SpO2 drop (Rule 1A)
    "min_spo2":           89.3,
    "confidence":         0.82,
    "rule_index":         1,              # which classification rule fired
    "flow_reduction":     0.45,           # fraction of baseline
    "classify_detail": {
        "effort_ratio":     0.72,
        "raw_var_ratio":    0.85,
        "phase_angle_deg":  52.3,
        "decision_reason":  "paradox_movement",
        "flattening_index": 0.28,
    },
    "spo2_cross_contaminated": False,
    "csr_flagged":        False,
}
\end{lstlisting}

\subsection{Summary Dictionary}

\begin{lstlisting}
{
    "ahi_total":       29.1,
    "oahi":            28.5,
    "cahi":            0.6,
    "n_obstructive":   85,
    "n_central":       2,
    "n_mixed":         1,
    "n_hypopnea":      78,
    "n_csr_flagged":   0,
    "n_local_baseline_rejected": 42,
    "severity":        "moderate",
}
\end{lstlisting}

% ================================================================
\section{Practical Tutorial: Processing an EDF File}
% ================================================================

A minimal end-to-end example:

\begin{lstlisting}
import mne
from psgscoring.pipeline import run_pneumo_analysis
from psgscoring.utils import build_sleep_mask

# 1. Load EDF
raw = mne.io.read_raw_edf("patient.edf", preload=True)

# 2. Assume hypnogram is available (from YASA or manual)
hypno = ["W","W","N1","N2","N2","N3", ...]  # per 30-s epoch

# 3. Define channel mapping
channel_map = {
    "flow":           "Pressure",      # nasal pressure
    "flow_thermistor":"Flow",           # oronasal thermistor
    "thorax":         "RIP Thora",
    "abdomen":        "RIP Abdom",
    "spo2":           "SpO2",
    "snore":          "Snore",
    "position":       "Pos.",
    "ecg":            "ECG II",
    "plm_l":          "PLMl",
    "plm_r":          "PLMr",
}

# 4. Run analysis
results = run_pneumo_analysis(
    raw, hypno, channel_map,
    scoring_profile="standard",
)

# 5. Extract results
resp = results["respiratory"]
print(f"AHI:  {resp['summary']['ahi_total']:.1f}/h")
print(f"OAHI: {resp['summary']['oahi']:.1f}/h")
print(f"Events: {len(resp['events'])}")

for ev in resp["events"][:5]:
    print(f"  {ev['type']:12s} at {ev['onset_s']:.0f}s, "
          f"dur={ev['duration_s']:.0f}s, "
          f"conf={ev['confidence']:.2f}")
\end{lstlisting}

% ================================================================
\section{Design Decisions and Trade-offs}
% ================================================================

\subsection{Why Rules, Not Deep Learning?}

\texttt{psgscoring} is deliberately rule-based:
\begin{enumerate}[nosep]
  \item \textbf{Transparency}: Every event can be traced to a specific
        AASM rule and threshold. A clinician can understand \emph{why}
        an event was scored.
  \item \textbf{Data efficiency}: Rules require zero training data.
        Deep learning requires thousands of expert-scored PSGs.
  \item \textbf{Auditability}: For clinical deployment (CE marking,
        FDA clearance), regulators require explainable decisions.
  \item \textbf{Correction specificity}: Each over/under-counting
        mechanism is independently addressable with its own counter.
\end{enumerate}

The planned hybrid approach uses a LightGBM confidence layer
\emph{on top of} the rule-based engine --- not replacing it.

\subsection{Why Not Use a Single Threshold?}

A single amplitude threshold (e.g., ``score hypopnea when envelope
$<$70\% of baseline'') fails because:
\begin{itemize}[nosep]
  \item The baseline itself is corrupted (Fixes 1, 6)
  \item The envelope misses per-breath reductions (Fix 7)
  \item SpO$_2$ coupling has timing issues (Fixes 2, 8)
  \item Signal artefacts generate false events (Fix 5)
\end{itemize}

Each correction addresses a specific failure mode.

\subsection{Performance Characteristics}

\begin{table}[H]
\centering\small
\begin{tabular}{@{}lr@{}}
\toprule
\textbf{Metric} & \textbf{Value} \\
\midrule
Processing time (8h PSG, 256 Hz) & 3--5 min \\
Peak RAM usage & $\sim$2 GB \\
Event detection (apnea + hypopnea) & $\sim$70 s \\
All 12 corrections overhead & $<$1 s \\
\bottomrule
\end{tabular}
\end{table}

% ================================================================
% (References moved to end of handbook)

% ================================================================
\section{Breath Detection and Flattening Index}
% ================================================================

Individual breath detection is essential for both peak-based hypopnea
detection (Fix~7) and RERA assessment (Fix~12).

\subsection{Breath Segmentation}

Breaths are detected by finding zero-crossings in the filtered flow
signal. Each breath consists of an inspiratory phase (positive flow)
and an expiratory phase (negative flow):

\begin{lstlisting}
from psgscoring.breath import detect_breaths

breaths = detect_breaths(filtered_flow, sf=256)
# Returns list of dicts:
# [{"onset_s": 10.2, "duration_s": 3.8,
#   "amplitude": 85.3, "insp_duration": 1.6}, ...]
\end{lstlisting}

\subsection{Flattening Index Computation}

The flattening index quantifies inspiratory flow limitation. A normal
breath has a triangular inspiratory waveform; a flattened breath
indicates upper airway resistance:

\begin{lstlisting}
from psgscoring.breath import compute_flattening_index

fi = compute_flattening_index(inspiratory_segment)
# fi ~ 0.10: normal (triangular)
# fi > 0.30: flattened (flow limitation)
# fi > 0.50: severely flattened
\end{lstlisting}

The index is defined as the fraction of inspiratory time during which
the signal exceeds 80\% of peak amplitude. Geometrically, this
measures how ``rectangular'' (plateau-shaped) the inspiratory waveform
is:

\begin{center}
\begin{tikzpicture}[scale=0.8]
  % Normal breath
  \draw[->] (0,0) -- (3.5,0) node[right] {\tiny t};
  \draw[->] (0,0) -- (0,2.5) node[above] {\tiny flow};
  \draw[thick,blue] (0,0) -- (1.5,2) -- (3,0);
  \draw[dashed,red] (0,1.6) -- (3,1.6) node[right] {\tiny 80\%};
  \node[below] at (1.5,-0.3) {\small Normal ($FI \approx 0.10$)};

  % Flattened breath
  \begin{scope}[xshift=5cm]
  \draw[->] (0,0) -- (3.5,0) node[right] {\tiny t};
  \draw[->] (0,0) -- (0,2.5) node[above] {\tiny flow};
  \draw[thick,blue] (0,0) -- (0.5,2) -- (2.5,2) -- (3,0);
  \draw[dashed,red] (0,1.6) -- (3,1.6) node[right] {\tiny 80\%};
  \fill[red,opacity=0.15] (0.4,1.6) rectangle (2.6,2);
  \node[below] at (1.5,-0.3) {\small Flattened ($FI \approx 0.45$)};
  \end{scope}
\end{tikzpicture}
\end{center}

\subsection{Breath Amplitude for Hypopnea Detection}

Per-breath amplitude is compared to a local baseline (median of
preceding 10~breaths):

\begin{lstlisting}
from psgscoring.breath import compute_breath_amplitudes

ratios = compute_breath_amplitudes(breaths, sf=256)
# ratios[i] = breath_amplitude[i] / median(prev 10 breaths)
# ratio < 0.70 -> this breath is part of a hypopnea
\end{lstlisting}

% ================================================================
\section{Periodic Limb Movement (PLM) Detection}
% ================================================================

PLMs are scored per AASM 2.6 and WASM (World Association of Sleep
Medicine) criteria:

\begin{itemize}[nosep]
  \item Duration: 0.5--10~seconds
  \item Amplitude: $\geq$8~$\mu$V above baseline (tibialis EMG)
  \item Inter-movement interval: 5--90~seconds
  \item A ``series'' requires $\geq$4 consecutive movements meeting
        the interval criterion
\end{itemize}

\begin{lstlisting}
from psgscoring.plm import detect_plm

result = detect_plm(plm_left, plm_right, sf=256,
                     hypno=hypno, sleep_mask=sleep_mask)
# result["summary"]["plmi"]  -> PLM index (/h sleep)
# result["summary"]["n_plm"] -> number of PLMs in series
\end{lstlisting}

The PLM index (PLMI) = PLMs in series / hours of sleep.
Clinical significance: PLMI $>$ 15/h suggests periodic limb
movement disorder (PLMD), especially if associated with arousals.

\begin{infobox}
\textbf{Respiratory association:} Leg movements occurring within 0.5~s
before to 0.5~s after a respiratory event are considered
respiratory-associated and are \emph{excluded} from the PLMI count.
This prevents respiratory events from inflating the PLM index.
\end{infobox}

% ================================================================
\section{Signal Quality Assessment}
% ================================================================

Automated scoring is only as reliable as the input signals. The
\texttt{signal\_quality} module evaluates every channel on four
dimensions:

\begin{table}[H]
\centering\small
\begin{tabular}{@{}llp{6cm}@{}}
\toprule
\textbf{Check} & \textbf{Threshold} & \textbf{Method} \\
\midrule
Flat-line & $>$10\% of recording & Rolling 2~s SD below role-specific
  threshold (0.5~$\mu$V for EEG, 0.1~mV for respiratory) \\
Clipping & $>$1\% of samples & Signal at ADC rail for $>$10
  consecutive samples \\
Line noise & $>$50\% of power & 49--51~Hz and 59--61~Hz power as
  fraction of total 1--Nyquist power \\
Disconnect & Any occurrence & Transitions from active signal to
  flat-line lasting $\geq$10~s \\
\bottomrule
\end{tabular}
\end{table}

Each channel receives a grade: \textbf{good}, \textbf{acceptable},
\textbf{poor}, or \textbf{unusable}.

\subsection{Montage Plausibility}

Cross-correlation checks detect hardware errors:
\begin{itemize}[nosep]
  \item EEG $\leftrightarrow$ EOG correlation $>$0.95: possible
        shared-reference or channel duplication
  \item Thorax $\leftrightarrow$ Abdomen correlation $>$0.98:
        possible duplication (prevents obstructive/central distinction)
\end{itemize}

% ================================================================
\section{Common Pitfalls for Developers}
% ================================================================

\subsection{Variable Scoping}

Local variables computed in outer functions must be passed explicitly
as parameters to inner functions. In Python, closures capture
variables by reference, not by value. The original codebase had a bug
where \texttt{\_tecg}, \texttt{\_r\_peaks}, and \texttt{\_sf\_ecg\_local}
were local variables in \texttt{detect\_respiratory\_events()} but
referenced inside \texttt{\_detect\_apneas()} --- causing
\texttt{NameError} in certain code paths.

\subsection{Docker Caching}

\texttt{docker compose restart} does \emph{not} rebuild images.
After modifying Python files, always use:
\begin{lstlisting}[language=bash]
docker compose build --no-cache
docker compose up -d
\end{lstlisting}

\subsection{Shell Escaping}

Bash interprets \texttt{\$} in bcrypt hashes. When writing passwords
to databases, use Python's \texttt{sqlite3} with parameterised queries:
\begin{lstlisting}
cursor.execute("UPDATE users SET password=? WHERE id=?",
               (hashed_pw, user_id))
\end{lstlisting}

\subsection{EDF Limitations}

\begin{itemize}[nosep]
  \item \texttt{pyedflib} is the reliable EDF+ writer;
        \texttt{edfio} silently writes plain EDF headers and drops
        annotations.
  \item MNE's EDF export fails on large physical min/max values.
  \item SomnoMedics stores position as ADC values (0--255), not
        integer codes --- requires auto-mapping.
\end{itemize}

\subsection{Translation Function Shadowing}

Never name local variables \texttt{t} when \texttt{t()} is imported
as a translation function:
\begin{lstlisting}
from i18n import t
# ...
t = date.today()   # BREAKS ALL t() CALLS!
today = date.today() # correct
\end{lstlisting}

% ================================================================
\section{Glossary}
% ================================================================

\begin{table}[H]
\centering\small
\begin{tabularx}{\textwidth}{@{}lX@{}}
\toprule
\textbf{Term} & \textbf{Definition} \\
\midrule
AHI & Apnea-Hypopnea Index: (apneas + hypopneas) / hours of sleep \\
OAHI & Obstructive AHI: only obstructive + mixed events \\
CAI & Central Apnea Index \\
RDI & Respiratory Disturbance Index: AHI + RERA index \\
RERA & Respiratory Effort-Related Arousal: flow limitation + arousal, no desaturation \\
ODI & Oxygen Desaturation Index: desaturations $\geq$3\% per hour \\
PLMI & Periodic Limb Movement Index: PLMs in series per hour of sleep \\
TST & Total Sleep Time (minutes of scored sleep) \\
TIB & Time In Bed \\
SE & Sleep Efficiency: TST / TIB $\times$ 100\% \\
SOL & Sleep Onset Latency: time from lights-off to first sleep epoch \\
WASO & Wake After Sleep Onset \\
RIP & Respiratory Inductance Plethysmography (effort bands) \\
TECG & Transformed ECG: high-pass filtered + QRS-blanked ECG \\
CSR & Cheyne-Stokes Respiration \\
MMSD & Mean Magnitude of Second Derivative \\
FRI & Flow Reduction Index: events with flow reduction but no AASM criteria met \\
EDF & European Data Format (PSG file format) \\
AASM & American Academy of Sleep Medicine \\
WASM & World Association of Sleep Medicine \\
\bottomrule
\end{tabularx}
\end{table}

% ================================================================
\section{Exercises}
% ================================================================

\begin{enumerate}
  \item \textbf{Signal processing:} Generate a synthetic nasal pressure
        signal (12 breaths/min, amplitude 100). Apply square-root
        linearisation. Compare the envelope before and after
        linearisation. What happens to the breath amplitude distribution?

  \item \textbf{Baseline sensitivity:} Create a 10-minute signal with
        stable breathing for 5~minutes, then a 30-second apnea followed
        by 4.5~minutes of recovery breathing at 150\% amplitude. Plot
        the dynamic baseline. Where does it become inflated? Implement
        the 30-second recovery mask (Fix~1) and compare.

  \item \textbf{Classification:} Create test inputs for
        \texttt{classify\_apnea\_type()} that trigger each of the 7
        rules. Verify that your inputs produce the expected type and
        confidence range.

  \item \textbf{Property testing:} Write a Hypothesis test that
        generates random \texttt{effort\_ratio} (0--2) and
        \texttt{raw\_var\_ratio} (0--3) values. Verify that the
        classifier never crashes and always returns a valid type.

  \item \textbf{SpO$_2$ coupling:} Generate a SpO$_2$ signal with a
        known 5\% desaturation at $t=40$~s. Test
        \texttt{get\_desaturation()} with window sizes of 30~s and
        45~s. At what circulatory delay does the 30~s window fail?

  \item \textbf{Flattening index:} Generate a triangular inspiratory
        waveform and a rectangular one. Compute the flattening index
        for each. What threshold best separates normal from flattened?

  \item \textbf{End-to-end:} Download a public PSG from PhysioNet
        (Sleep-EDF dataset). Run YASA staging, then
        \texttt{run\_pneumo\_analysis()} with all three scoring profiles.
        Compare the AHI across profiles and explain the differences.
\end{enumerate}

% ================================================================
\section*{Acknowledgments}

This handbook accompanies the \texttt{psgscoring} library, which
extends YASA by Raphael Vallat and Matthew P.\ Walker. The staging
classifier was trained on SHHS and MESA data funded by the NHLBI.

\vspace{1cm}
\begin{center}
\texttt{pip install psgscoring}\\[4pt]
\url{https://github.com/bartromb/psgscoring}\\[4pt]
\url{https://slaapkliniek.be}
\end{center}

% ================================================================
\newpage
\section{References}
\label{sec:references}
% ================================================================

\subsection*{Standards and Scoring Manuals}

\begin{enumerate}[nosep,label={[\arabic*]}]
  \item Berry RB, Brooks R, Gamaldo C, et~al. \emph{The AASM Manual for
        the Scoring of Sleep and Associated Events: Rules, Terminology and
        Technical Specifications. Version~2.6.} American Academy of Sleep
        Medicine, 2020.
  \item Iber C, Ancoli-Israel S, Chesson AL, Quan SF. \emph{The AASM
        Manual for the Scoring of Sleep and Associated Events: Rules,
        Terminology and Technical Specifications.} 1st~ed. AASM, 2007.
  \item Zucconi M, Ferri R, Allen R, et~al. The official World Association
        of Sleep Medicine (WASM) standards for recording and scoring periodic
        leg movements in sleep (PLMS) and wakefulness (PLMW) developed in
        collaboration with a task force from the International Restless Legs
        Syndrome Study Group (IRLSSG). \emph{Sleep Med}.
        2006;7(2):175--183.
        \href{https://doi.org/10.1016/j.sleep.2006.01.001}{doi:10.1016/j.sleep.2006.01.001}.
\end{enumerate}

\subsection*{Sleep Staging and YASA}

\begin{enumerate}[nosep,label={[\arabic*]},resume]
  \item Vallat R, Walker MP. An open-source, high-performance tool for
        automated sleep staging. \emph{eLife}. 2021;10:e70092.
        \href{https://doi.org/10.7554/eLife.70092}{doi:10.7554/eLife.70092}.
  \item Perslev M, Darkner S, Kempfner L, Nikolic M, Jennum PJ, Igel C.
        U-Sleep: resilient high-frequency sleep staging. \emph{npj Digital
        Medicine}. 2021;4:72.
        \href{https://doi.org/10.1038/s41746-021-00440-5}{doi:10.1038/s41746-021-00440-5}.
  \item Ke G, Meng Q, Finley T, Wang T, Chen W, Ma W, Ye Q, Liu T-Y.
        LightGBM: a highly efficient gradient boosting decision tree.
        \emph{Advances in Neural Information Processing Systems}.
        2017;30:3146--3154.
        \href{https://proceedings.neurips.cc/paper/2017/hash/6449f44a102fde848669bdd9eb6b76fa-Abstract.html}{NeurIPS~2017}.
  \item Gramfort A, Luessi M, Larson E, et~al. MNE software for processing
        MEG and EEG data. \emph{NeuroImage}. 2014;86:446--460.
        \href{https://doi.org/10.1016/j.neuroimage.2013.10.027}{doi:10.1016/j.neuroimage.2013.10.027}.
  \item Feinberg I, Floyd TC. Systematic trends across the night in human
        sleep cycles. \emph{Psychophysiology}. 1979;16(3):283--291.
        \href{https://doi.org/10.1111/j.1469-8986.1979.tb02991.x}{doi:10.1111/j.1469-8986.1979.tb02991.x}.
\end{enumerate}

\subsection*{Respiratory Scoring and Signal Processing}

\begin{enumerate}[nosep,label={[\arabic*]},resume]
  \item Thurnheer R, Xie X, Bloch KE. Accuracy of nasal cannula pressure
        recordings for assessment of ventilation during sleep. \emph{Am J
        Respir Crit Care Med}. 2001;164(10):1914--1919.
        \href{https://doi.org/10.1164/ajrccm.164.10.2010113}{doi:10.1164/ajrccm.164.10.2010113}.
  \item Montserrat JM, Farr\'e R, Ballester E, Felez MA, Past\'o M,
        Navajas D. Evaluation of nasal prongs for estimating nasal flow.
        \emph{Am J Respir Crit Care Med}. 1997;155(1):211--215.
        \href{https://doi.org/10.1164/ajrccm.155.1.9001310}{doi:10.1164/ajrccm.155.1.9001310}.
  \item Lee H, Park J, Kim H, Lee KJ, Heo C. Detection of apneic events
        from single-channel nasal airflow using 2nd derivative method.
        \emph{Physiol Meas}. 2008;29:N37--N45.
        \href{https://doi.org/10.1088/0967-3334/29/5/N01}{doi:10.1088/0967-3334/29/5/N01}.
  \item Hosselet J, Norman RG, Ayappa I, Rapoport DM. Detection of flow
        limitation with a nasal cannula/pressure transducer system. \emph{Am
        J Respir Crit Care Med}. 1998;157(5):1461--1467.
        \href{https://doi.org/10.1164/ajrccm.157.5.9708008}{doi:10.1164/ajrccm.157.5.9708008}.
  \item Uddin MB, Chow CM, Su SW. A novel algorithm for automatic
        diagnosis of sleep apnea from airflow and oximetry signals.
        \emph{Physiol Meas}. 2021;42:015001.
        \href{https://doi.org/10.1088/1361-6579/abd47a}{doi:10.1088/1361-6579/abd47a}.
  \item Nakano H, Tanigawa T, Furukawa T, Nishima S. New rule-based
        algorithm for real-time detecting sleep apnea events. \emph{J Clin
        Sleep Med}. 2016;12(10):1389--1396.
        \href{https://doi.org/10.5664/jcsm.6200}{doi:10.5664/jcsm.6200}.
\end{enumerate}

\subsection*{Apnea Classification and Effort Assessment}

\begin{enumerate}[nosep,label={[\arabic*]},resume]
  \item Berry RB, Albertario CL, Guttman B, et~al. Use of a transformed
        ECG signal to detect respiratory effort during apnea. \emph{J Clin
        Sleep Med}. 2019;15(11):1653--1660.
        \href{https://doi.org/10.5664/jcsm.7880}{doi:10.5664/jcsm.7880}.
  \item Berry RB, Budhiraja R, Guttman B, et~al. Use of chest wall
        electromyography to detect respiratory effort during
        polysomnography. \emph{J Clin Sleep Med}. 2016;12(9):1239--1244.
        \href{https://doi.org/10.5664/jcsm.6122}{doi:10.5664/jcsm.6122}.
\end{enumerate}

\subsection*{Scoring Variability and Validation}

\begin{enumerate}[nosep,label={[\arabic*]},resume]
  \item Rosenberg RS, Van Hout S. The American Academy of Sleep Medicine
        inter-scorer reliability program: sleep stage scoring. \emph{J Clin
        Sleep Med}. 2013;9(1):81--87.
        \href{https://doi.org/10.5664/jcsm.2350}{doi:10.5664/jcsm.2350}.
  \item Rosenberg RS, Hirshkowitz M, Engleman HM, Penzel T. The AASM
        inter-scorer reliability program: respiratory events. \emph{J Clin
        Sleep Med}. 2014;10(4):447--454.
        \href{https://doi.org/10.5664/jcsm.3630}{doi:10.5664/jcsm.3630}.
  \item Ruehland WR, Rochford PD, O'Donoghue FJ, Pierce RJ, Singh P,
        Thornton AT. The new AASM criteria for scoring hypopneas: impact on
        the apnea hypopnea index. \emph{Sleep}. 2009;32(2):150--157.
        \href{https://doi.org/10.1093/sleep/32.2.150}{doi:10.1093/sleep/32.2.150}.
  \item Malhotra A, Younes M, Kuna ST, Benca R, Kushida CA, Walsh J,
        Hanlon A, Staley B, Pack AI, Pien GW. Performance of an automated
        polysomnography scoring system versus computer-assisted manual
        scoring. \emph{Sleep}. 2013;36(4):573--582.
        \href{https://doi.org/10.5665/sleep.2548}{doi:10.5665/sleep.2548}.
\end{enumerate}

\subsection*{Novel Metrics and Complementary Approaches}

\begin{enumerate}[nosep,label={[\arabic*]},resume]
  \item Parekh A, Kam K, Wickramaratne S, et~al. Ventilatory burden as a
        measure of obstructive sleep apnea severity is predictive of
        cardiovascular and all-cause mortality. \emph{Am J Respir Crit Care
        Med}. 2023;208(11):1216--1226.
        \href{https://doi.org/10.1164/rccm.202301-0109OC}{doi:10.1164/rccm.202301-0109OC}.
\end{enumerate}

\subsection*{Epidemiology}

\begin{enumerate}[nosep,label={[\arabic*]},resume]
  \item Peppard PE, Young T, Barnet JH, Palta M, Hagen EW, Hla KM.
        Increased prevalence of sleep-disordered breathing in adults.
        \emph{Am J Epidemiol}. 2013;177(9):1006--1014.
        \href{https://doi.org/10.1093/aje/kws342}{doi:10.1093/aje/kws342}.
  \item Benjafield AV, Ayas NT, Eastwood PR, et~al. Estimation of the
        global prevalence and burden of obstructive sleep apnoea: a
        literature-based analysis. \emph{Lancet Respir Med}.
        2019;7(8):687--698.
        \href{https://doi.org/10.1016/S2213-2600(19)30198-5}{doi:10.1016/S2213-2600(19)30198-5}.
\end{enumerate}

\subsection*{Software and Statistical Tools}

\begin{enumerate}[nosep,label={[\arabic*]},resume]
  \item Vallat R. Pingouin: statistics in Python. \emph{J Open Source
        Software}. 2018;3(31):1026.
        \href{https://doi.org/10.21105/joss.01026}{doi:10.21105/joss.01026}.
\end{enumerate}

\end{document}
